Six ways to understand monads\\ \small{Experiences from the functional programming course at TU Delft}

### 2012-11-01 latex/paper.tex
Erik's corrections

Due to a variety of reasons, including the advent of cloud computing, the rising
rate of information production and the necessity to reach the market fast,
currently, large corporations and start-ups alike are investigating alternative
programming and information storage models. As a result, during the last few
years, the practical software engineering field is witnessing a noticeable shift
towards functional programming. Scripting languages, notably Javascript and
Ruby, pioneered the introduction of functional concepts, such as closures and
lambda functions, to mainstream programming. A new wave of programming
languages, developed to overcome the expressiveness and complexity limitations
exhibited in mainstream languages, have promoted functional constructs, such as
type safe pattern matching, higher order functions and single assignment
variables, to first class citizens (Scala).  New, purely functional, languages
have emerged to fill in the remaining gaps (F\#, Clojure), often introducing
significant advancements in their field of specialisation (such as Erlang in
distributed fault-tolerant systems).  Finally, large scale information
processing systems such as Map/Reduce~\cite{Dean04} and domain specific
languages such as {\sc linq} have integrated functional concepts to ease the
expression of computations.

@@

Functional programming is arguably the oldest programming paradigm. In it is
purest form, it is based on a minimalistic theoretical background
($\lambda$-calculus~\cite{Baren84}). Relatively recent
formulations~\cite{Meije91, Wadle93} also introduced concepts from category
theory. While the every day use of functional programming does not necessarily
require the programmer to be aware of the theory, understanding it usually leads
to more elegant algorithmic solutions. Functional programming teaching is also
associated with advanced type systems; indeed, the flagship functional
programming languages (Haskell and the {\sc ml} family) both offer very
sophisticated support for types. Consequently, teaching the full spectrum of
functional programming theory and techniques would have been impossible in a
half-semester course; instead we decided to focus on practical aspects of data
processing, transformations and state representation using functional
techniques.

@@

Programming languages, in addition to enabling the programmer to express a
series of instructions to be executed by a computer, affect the programmer's
thought process and consequently her approach towards problem
solving~\cite{Ivers80}.  When teaching a programming course, it is important to
be able to demonstrate concepts without interference by the chosen language's
syntax. However, a non-practical language might have the opposite effect; our
experience has shown that the further a demonstration language is from practical
application, the less important student feel the taught concepts are.
Fortunately, most functional programming concepts can be expressed cleanly by
several widespread languages: for example, Javascript, Ruby and {\sc c\#} have
closures, first class and higher order functions, while a number of emerging
languages, such as Scala, {\sc f\#} and Clojure are functional languages
implemented on familiar development platforms. Apart from staying compatible
with the existing literature, there is no practical reason that necessitates
teaching of functional programming strictly in a language like Haskell or {\sc
ml}.

@@

The course consisted of a series of lectures, of which two where invited
lectures by an external instructor, and laboratory sessions. In total, the
course consisted of 14 hours of lectures and 8 hours of lab work.
The students
also had to carry out a series of homework assignments and a final project.
The students would be evaluated by their performance on the final assignment. 
The lectures took place in a quarter (half semester), during 3 intense 
weeks; the students had another 6 weeks to finish their final projects.

@@

  \item Functions and functional composition, higher order functions, recursive
    functions, avoiding recursion

  \item Basic types, function types, currying

  \item Lists, mapping and traversal using folds

  \item Monads, composition and their application on the {\sc linq} query 
    language~\cite{Meije11}

  \item Monoids, functors and their application on data structure processing

  \item Functional formalization of the Map/Reduce data processing
    paradigm~\cite{Lamme08}

@@

  \item Functional programming works best in data transformation scenarios

@@

The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
graphics primitive of the student's choice using only right triangles.  In its
core, the exercise required students to decompose arbitrary triangles, which
comprised the input Utah Teapot model, to a series right triangles. The students
had to come up with the decomposition method (using an analytic geometry
method), a decomposition termination criterion to stop the decomposition when
triangles are too small to be rendered on screen and a method to recursively
apply the above mentioned transformations on the input data. As always, the
students could decide the implementation language of their choice. The exercise
was given as a mid-week homework assignment, between the Monday and the Friday
lectures.

>>>>

Due to a variety of reasons, including the advent of cloud computing, the rising
rate of information production and the necessity to reach the market fast,
currently, large corporations and start-ups alike are investigating alternative
programming and information storage models. As a result, during the last few
years, the practical software engineering field is witnessing a noticeable shift
towards functional programming. Scripting languages, notably Javascript and
Ruby, pioneered the introduction of functional concepts, such as closures and
lambda functions, to mainstream programming. A new wave of programming
languages, developed to overcome the expressiveness and complexity limitations
exhibited in mainstream languages, have promoted functional constructs, such as
type safe pattern matching, higher order functions and single assignment
variables, to first class citizens (Scala, {\sc c\#}).  New, functional
languages have emerged to fill in the remaining gaps (F\#, Clojure, Racket),
often introducing significant advancements in their field of specialisation
(such as Erlang in distributed fault-tolerant systems).  Finally, large scale
information processing systems such as Map/Reduce~\cite{Dean04} and domain
specific languages such as {\sc linq} have integrated functional concepts to
ease the expression of masively parallel computations.

@@

Functional programming is arguably one of the oldest programming paradigms. In it is
purest form, it is based on a minimalistic theoretical background
($\lambda$-calculus~\cite{Baren84}). Relatively recent
formulations~\cite{Meije91, Wadle93} also introduced concepts from category
theory. While the every day use of functional programming does not necessarily
require the programmer to be aware of the theory, understanding it usually leads
to more elegant algorithmic solutions. Functional programming teaching is also
associated with advanced type systems; indeed, the flagship functional
programming languages (Haskell and the {\sc ml} family) both offer very
sophisticated support for types. Consequently, teaching the full spectrum of
functional programming theory and techniques would have been impossible in a
half-semester course; instead we decided to focus on practical aspects of data
processing, transformations and state representation using functional
techniques.

@@

Programming languages, in addition to enabling the programmer to express a
series of instructions to be executed by a computer, affect the programmer's
thought process and consequently her approach towards problem
solving~\cite{Ivers80}.  When teaching a programming course, it is important to
be able to demonstrate concepts without interference by the chosen language's
syntax. However, a non-practical language might have the opposite effect; our
experience has shown that the further a demonstration language is from practical
application, the less important students feel the taught concepts are.
Fortunately, most functional programming concepts can be expressed cleanly by
several widespread languages: for example, Javascript, Ruby and {\sc c\#} have
closures, first class and higher order functions, while a number of emerging
languages, such as Scala, Racket, {\sc f\#} and Clojure are functional languages
implemented on familiar development platforms. Apart from staying compatible
with the existing literature, there is no practical reason that necessitates
teaching strictly in a language like Haskell or {\sc ml}.

@@

The course consisted of a series of lectures, of which two where invited
lectures by an external instructor, and laboratory sessions. In total, the
course consisted of 14 hours of lectures and 8 hours of lab work.
The students
also had to carry out a series of homework assignments and a final project.
The evaluation would be based on student performance on the final assignment. 
The lectures took place in a quarter (half semester), during 3 intense 
weeks; the students had another 6 weeks to finish their final projects.

@@

  \item Functions and functional composition, higher order functions, recursion, 
    avoiding recursion via higher order functions.

  \item Basic types, function types, currying.

  \item Lists, mapping and traversal using folds.

  \item Monads, composition and their application on the {\sc linq} query 
    language~\cite{Meije11}.

  \item Monoids, functors and their application on data structure processing.

  \item Functional formalization of the Map/Reduce data processing
    paradigm~\cite{Lamme08}.

@@

  \item Functional programming works best in data transformation scenarios.

  \item Complex problems can be solved by decomposing them into smaller and simpler ones.

@@

The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
graphics primitive of the student's choice using only right triangles (with one
horizontal leg).  In its
core, the exercise required students to decompose a collection of arbitrary triangles, which
comprised the input Utah Teapot model, to a collection of right triangles. The students
had to come up with the decomposition method (using some simple analytic
geometry, this was not a course on 3D graphics), a decomposition termination criterion to stop the decomposition when
triangles were too small to be rendered on screen and a method to recursively
apply the above mentioned transformations on the input data. As always, the
students could decide the implementation language of their choice. The exercise
was given as a mid-week homework assignment, between the Monday and the Friday
lectures.
